### 2020-02-27 docs/papers/osr/body.tex
a draft

Systems verification is picking up momentum.
The promise is that we might someday replace a methodology of
write-test-debug with a comprehensive compile-time confirmation
of correctness. In particular, critical software could remove
``ship'' from the bug-fix development cycle.

>>>>

For half a century, we have known that a program is constructed by
writing code, testing, and debugging. Program verification promises
an alternative process in which no bug escapes the compiler.

Convincing a mechanical checker that a program is correct requires
resolving a fantastic degree of detail.
The SMART system offered an extension to Paxos~\cite{smart}.
Its authors provided a hand-written proof at a level of detail intended
to satisfy a mechanical checker, although it was never checked.
Just the proof of the SMART algorithm, without an imperative implementation,
is 70 pages of text~\cite{smart-proof}.
If we are to apply verification as a practical engineering tool,
we must rely heavily on automation to retire tedious logic obligations,
focusing humans on just the subtle invariants that reflect the insights
of the system design.

The authors have verified several systems using highly-automated
verification tools. This paper conveys the experience of learning and using
such tools. We argue that strong automation is the path forward for
large-scale systems verification, and we highlight open challenges.

\section{A verified codebase}

The code base of a verified system has three parts:
the specification, the implementation, and the proof.

The \textit{specification} defines the application spec
(what does the program do?) and the environment spec
(what can the program assume about its environment?).

The \textit{implementation} is the source code that will compile
into the final program. It is imperative (has in-place updates) and
uses machine-supported datatypes (uint64) to give the programmer
enough control to achieve great performance.

The \textit{proof} files illuminate the invariants and structure
the arguments that show that the implementation does what the specification
says.

A typical specification is smaller than 1/10^{th} the size of
the implementation it constrains.
In {\ironstar}, the spec files carry a distinguished file extension,
and spec files may only include other spec files.
If the developer reads all of the spec files, she need read no other
files to understand what the verified program does.

An {\ironstar} proof text is around five times
as big as the implementation code.
The effort spent constructing and maintaining it is in lieu of
maintaining a conventional test suite.

\section{A day in the life of a verification engineer}

At the large scale, the team's task is to write a fast implementation,
and show that its behavior matches the spec. This task breaks down into
making smaller arguments; a typical argument might say ``splitting a node
in a tree creates a new tree that represents all the same key-value pairs.''

When using an automation-centric tools, the engineer begins each such
argument the same way: She states the argument, and then offers an empty
proof:
\begin{verbatim} % TODO get an environment that numbers lines.
lemma TODOPlausibleLemmaName(before:Tree, after:Tree, pivot:Key)
  requires TreeInv(a)
  requires SplitNodeRelation(before, after, pivot)
  ensures TreeInv(b)
{
}
\end{verbatim}
Then she asks the verifier to try to verify the lemma.
This is analogous to testing every possible argument assignment,
but of course the mechanism is not actually model checking, since
there are infinite such assignments.

Sometimes, automation
successfully opens the tall tree of definitions under
\texttt{SplitNodeRelation} and \texttt{TreeInv}, and confirms the validity
of the lemma.
More commonly, the verifier cannot convince itself, and it reports
\texttt{TODO real text at line 4}.
Maybe the lemma is invalid: The definitions don't actually
do what the engineer hoped, either because of a deep misunderstanding
or due to a minor off-by-one error.
This is why we're using automation in the first place; it's a failed test
pointing to a flaw in the designer's understanding.
Or perhaps the lemma is valid, but the automation can't see why.
This is by far the more common case; it's the tax we pay to enjoy
exhaustive testing.

However, the engineer's action in either case is the same: she must
elaborate the proof until either the verifier admits its validity
or she discovers an underlying error.

The currency of definitions is are quantified logical statments.
\verb{TreeInv(t)} is a list of conjuncts; one says that the tree
is acyclic:
\begin{verbatim}
predicate Acyclic(t) {
  forall p, i, j :: Path(p, t) && i!=j ==> p[i] != p[i]
}
\end{verbatim}
This says that no path through the tree visits the same node twice.

Suppose the engineer wonders if acyclicity is the thing tripping up the
verifier. So she replaces the lemma proof with:
\begin{verbatim}
{
  assert Acyclic(b);
}
\end{verbatim}
The verifier will respond in one of the following ways:
\begin{itemize}
  \item line 6 assertion failed: Yep, acyclicity is the difficult part,
  \textit{and if we assume it's true, the postcondition verifies}.
  \item line 4 postcondition failed: acyclicity was valid; some other
  conjunct in TreeInv(b) is the problem.
  \item \textit{both} errors: The verified needs help seeing more than one
  conjunct in \texttt{TreeInv}.
  \item neither error: the assertion itself has given the verifier a
  sufficient hint to verify the lemma (see \S\ref{observes}).
  \item \textit{Timeout}. See \S\ref{timeouts}.
\end{itemize}
The first two cases are by far the most common: the engineer has successfully
"bisected" the problem, either implicating or vindicating the \verb{Acyclic}
conjunct.

She applies this bisection technique recursively, just as she would use
printf bisection debugging to back a deterministic runtime test failure into a
corner.
If \verb{Acyclic(b)} is unproven, she drills down:
\begin{verbatim}
{
  forall p, i, j | Path(p, t) && i!=j
    ensures p[i] != p[i]
  {
  }
  assert Acyclic(b);
}
\end{verbatim}
The forall brings new variables \verb{p, i, j} into scope for use in its
proof body which, again, begins as \verb{\{\}}.
The verifier complains about the postcondition on line 7.
It accepts the assert on line 10, predicated on the assumption that the
engineer will ultimately correct the failing forall statement.

Now that \verb{p, i, j} have been introduced, the engineer may drill
further down into definitions to argue why \verb{p[i] != p[i]}.
In this exmaple, she would begin a proof by contradiction:
If there exists a cyclic path \verb{pb} in \verb{b},
then construct a witness path \verb{pa} in tree \verb{a} that is
cyclic, which contradicts the given \verb{Acyclic(a)}.

This iterative, user-guided debugging approach stands in contrast to other
common verification development methodologies.
The engineer isn't confronted with a half-transformed proof goal
with a hundred subterms, forcing her to
think in the verifier's internal representation.
Nor does she dig through a counterexample with a value assigned for each
of hundreds of in-scope variables, only two of which are actually relevant.
Both of those approaches present information to the user that scales up with
the system size, and both ask the user to interpret the system's model.
By contrast, bisection proof debugging lets the engineer develop her own
mental model by probing the verification Oracle in a game of twenty-questions.
The benefit of letting the engineer steer the story and maintain her own
mental model outweighs the fact that she must make a logarithmic number of
inquiries to reach the root cause, especially because each inquiry is answered
rapidly.

Returning to the possible verifier responses: in case 3, the verifier
complains about both the assertion and the postcondition; the engineer
has failed to bisect the problem. It could be as simple as the assertion
being incorrect, or it could be that there is more than one problem
(error in definition, or missing proof) to identify. The engineer proceeds
the same as above, but has more ground to cover before she's done.

Case 4 is a victory: by introducing the assertion, the engineer has
triggered the automation to exploit a quantifier it hadn't considered
automatically, and it has completed the proof.

Case 5, a timeout, is a meta-problem: not only can the verifier not
see the proof, but the automation strategy has broken down, and requires
human intervention to fix. These issues are rare: we bisect-debug all day,
but have to resolve a timeout perhaps once a week.
We discuss automation control and timeout mitigation in Section~\ref{timeout}.

Other than timeout management, note that the nitty-gritty work of verification
engineering feels a lot like printf-debugging a deterministic test failure.
The process is familiar, easy to learn, and it generally requires deep insight
only about the system being debugged, not about the verification tool itself.

TODO the glory of quantified vs recursive definitions.

\subsection{Structuring the proof}

While the low-level mechanics are simple, the verification engineer does
need to adopt new disciplines to organize her proof text for scale and
maintainability. The goal is the same as organizing a big software system
into modules for maintainability, but the techniques for arranging a
proof argument are different.

A single line of an imperative program may be correct for many reasons,
at multiple levels of abstraction, all of which must hold to support the
program's correctness. For example, in this line:
\begin{verbatim}
  node->writeToBlock(lba)
\end{verbatim}
Dereferencing \verb{node} is safe because \verb{node} was allocated earlier
in the method and not freed. The write to address \verb{lba} is safe
because \verb{lba} was returned from a library function managing free space
on the device. The content written, a marshalled representation of \verb{node},
is appropriate because later in the program we'll store a pointer to \verb{lba}
that expects a representation of \verb{node} to appear there.

Common static analysis tools can confirm non-null properties like the first
one. The other arguments are at higher levels of abstraction, interact with
distant lines of code in the program, and may even interect with other
instances of the process, such as before or after a crash event that resets
RAM. In a conventional program, each informal argument might appear in a
comment on the code line, or might be inferrable from the name of
the methods or the methods it calls, or may appear in an Engilsh design
document from two years ago, or may appear nowhere but in an engineer's
memory.

Mechanical verification requires that each such argument be written down.
Trying to cram all of these arguments into the ground-level source code
is overwhelming; the more abstract arguments are weighed down with the
tedium of implementation. In {\ironstar}, we employ state machine refinement
to organize proof text%
\footnote{We contrast this use with other verification efforts that also
use refinement, but in service to the verification system, to divide the
proof up into layers so fine that automation can leap from one to the next
without assistance. In {\ironstar}, refinement layers are introduced at
a granularity that humans find natural, just as developers choose appropriate
granularity boundaries for functional abstraction or object encapsulation.
}
by separating it along the axis of abstraction.

% TODO discuss app specification in more detail somewhere?
The app spec is a TLA+-style state machine.
In {\veribetrfs}, the spec is a key-value dictionary extended with
a \verb{flush} operation and a \verb{crash} event that interleaves
at the whim of the environment. These specs are all written with immutable
data structures and mathematical conveniences (unbounded integers, infinite
sets and sequences) that let us focus on the behavior while ignoring
implementation details.

The first layer of refinement introduces the structure of a {\bepsilon}
tree, while abstracting the nodes as infinite sets mapping every key to
a child node.
Proof text at this layer argues that the tree state machine maintains the
necessary invariants such that the tree acts exactly like a dictionary.

The second layer refines the infinite nodes to ``pivot'' nodes, which
map keys from A---G to one child, H---N to the next, and so on. Proof
text at this layer argues that pivot nodes are a (finite!) shorthand
for infinite maps, and hence that {\bepsilon}{\langle}pivot{\rangle} is
still a dictionary.

The previous layer supposed a disk that accepts the Node abstract data
type as arguments to \verb{write}; the third layer refines the Pivot Node to
have a marshalled representation. Proof text at this layer argues
that \verb{unmarshall(marshall(n) == n}, and thus that, when
a {\bepsilon} tree made of marshallable nodes is connected to a disk
that does I/O on byte strings, the system behaves the same as the prior layer.

Finally, the bottom-most layer replaces immutable datatypes with mutable
ones (to exploit the performance of in-place updates), and replaces
mathematical integers with machine \verb{uint64}s. Proof text at this
layer shows that aliasing is controlled enough that mutable data structures
mimic their immutable models above. It also shows that bounds are respected
so that \verb{uint64}s don't overflow.

Each of the arguments is necessary to stitch the proof together from the bottom
layer implementation to the top level application spec.
The important observation from this tour is that the refinement layers
act as narrow interfaces along the axis of abstraction: arguments about
high-level concepts like {\bepsilon} tree structure may omit details
about marshalling and register bounds; conversely, low-level arguments
may attend only to detailed reasoning, as the abstract properties were
dispensed with elsewhere.

Module design in an imperative program creates conceptual boundaries,
so that an engineer can study and modify part of the program while being
insulated from interactions with most of the other lines of code in the
system.
Likewise, we have found this organization enables us to study and modify
parts of our verified system while insulating our changes from other
parts. Most importantly, it means that we can introduce a new feature
(like sub-node reads) while leaving most of the proof text undisturbed.
Because proof is 80\% of the text of our verified system,
incremental and modular maintenance is as important for the proof as
for the imperaive code.

------------------------------------------------------------------------------
Systems verification is picking up momentum.
The promise is that we might someday replace a methodology of
write-test-debug with a comprehensive compile-time confirmation
of correctness. In particular, critical software could remove
``ship'' from the bug-fix development cycle.
